Human capital mediates natural selection in contemporary humans

### 2022-05-03 why-natural-selection.Rmd
Tweak text

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].

@@

res_significance <- res_wt_combined %>% 
                      group_by(term) %>% 
                      summarize(
                        all = all(p.value < 0.05/33), 
                        any = any(p.value < 0.05/33)
                      )
res_sig_all <- sum(res_significance$all)
res_sig_any <- sum(res_significance$any)

@@

Figure \@ref(fig:plot-res-wt) plots selection effects among the entire
sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
participants using the model of @van2021effects. Weighting makes a
large difference to estimates, increasing effect sizes by a mean of 
`r percent(mean_rel$Main - 1)`.[^weighting]

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

We emphasize that our categories are not necessarily exogenous to polygenic
scores. For example -- both in the data (Appendix Figure
\@ref(fig:plot-age-flb-dv)) and in our model -- age at first live birth is a
choice variable, which is endogenous to an individual's human capital and
related polygenic scores. Nevertheless, observed differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible
theories; a good theory of contemporary natural selection needs to explain
how these results arise. As we show below, a model based on the economic
theory of fertility can do just that.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

>>>>

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias in the sample [@10.1093/aje/kwx246].

@@

n_sig <- sum(res_wt_van_alten_weights$p.value < 0.05/33)

@@

Figure \@ref(fig:plot-res-wt) plots selection effects in the entire
sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
participants using the model of @van2021effects. Weighting makes a
large difference: effect sizes go up by a mean of 
`r percent(mean_rel$Main - 1)`.[^weighting] `n_sig` out of 33 selection effects
are significant at $p$ < 0.05/33.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
Turning to family structure, we split respondents by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
Now, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

@@

We emphasize that our categories are not exogenous to polygenic scores. For
example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
our theoretical model -- age at first live birth is a choice variable, which is
endogenous to a person's human capital and their related polygenic scores.
Nevertheless, differences in selection effects across subgroups constrain the
set of possible theories; a good theory of contemporary natural selection needs
to explain how these results come about. As we show below, a model based on the
economic theory of fertility can do just that.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We don't find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).
